---
title: "PPCSexRx: From Evidence to Bedside Prescription"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PPCSexRx: From Evidence to Bedside Prescription}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(PPCSexRx)
```

## Background

Persistent post-concussion symptoms (PPCS) — defined as symptoms lasting 28 or
more days after injury — affect an estimated 15–30% of adolescent concussion
cases. Despite growing evidence supporting sub-symptom threshold aerobic exercise
(SSTAE) as a first-line rehabilitation strategy, most clinicians still rely on
static PDF protocols that cannot respond to individual patient data.

**PPCSexRx** translates the evidence synthesised in Li (2026) — winner of the
2026 NATA Foundation Best Summary Evidence Research Award — into a fully
interactive clinical decision support tool. The package implements three
functions that mirror the real clinical workflow:

```
screen_ppcs()  →  prescribe_ppcs()  →  track_progress()
```

> **GRADE notice:** All recommendations in this package reflect LOW certainty
> evidence with a conditional recommendation FOR SSTAE in adolescents with PPCS
> (Li, 2026). Clinical judgement remains essential.

---

## Case 1: Rural High School — No BCTT Equipment Available

A 16-year-old football player presents 35 days post-concussion with ongoing
headache and fatigue. The athletic trainer works at a rural high school with no
Buffalo Concussion Treadmill Test (BCTT) equipment.

### Step 1: Eligibility Screen

```{r case1-screen}
screen_ppcs(age = 16, days_post_injury = 35)
```

The athlete is eligible. No vestibular, cervical, or vision contraindications
are present.

### Step 2: Exercise Prescription (Age-Predicted Fallback)

Because no BCTT is available, `prescribe_ppcs()` automatically uses the
age-predicted heart rate method (60–70% HRmax), as described in Li (2026), p.11.

```{r case1-prescribe}
rx1 <- prescribe_ppcs(age = 16, days_post_injury = 35)
rx1
```

The prescription recommends **`r rx1$target_hr` bpm** for 20 minutes, 5 days
per week — immediately implementable with any HR monitor, no specialist
equipment required.

### Step 3: Track Progress Across Sessions

The athletic trainer records each session. After two sessions without worsening,
the protocol advances intensity by 5 bpm.

```{r case1-track}
# Session 1
t1 <- track_progress(
  log              = NULL,
  current_pcss     = 28,
  current_hr       = rx1$target_hr,
  current_duration = 18,
  prescription     = rx1
)

# Session 2 — slight improvement
t2 <- track_progress(
  log              = t1$updated_log,
  current_pcss     = 25,
  current_hr       = rx1$target_hr,
  current_duration = 20,
  prescription     = rx1
)

# Session 3 — continued improvement: progression triggered
t3 <- track_progress(
  log              = t2$updated_log,
  current_pcss     = 21,
  current_hr       = rx1$target_hr,
  current_duration = 20,
  prescription     = rx1
)
t3
```

After `r t3$sessions_total` sessions with no worsening, the recommended HR
advances to **`r t3$adjust_hr` bpm** — consistent with the progression rule
in Li (2026), p.14.

### Visualise the Rehabilitation Trajectory

```{r case1-plot, fig.width = 6, fig.height = 5}
plot(t3)
```

The dual-panel plot shows PCSS symptom burden (top) and achieved heart rate
(bottom) across sessions, giving the clinician an immediate visual of recovery
trajectory.

---

## Case 2: University Clinic — BCTT Available

A 17-year-old swimmer presents 40 days post-concussion. The sports medicine
clinic has BCTT equipment; the athlete's symptom threshold HR (HRST) was
measured at 155 bpm.

### Step 1: Screen

```{r case2-screen}
screen_ppcs(age = 17, days_post_injury = 40)
```

### Step 2: BCTT-Guided Prescription

```{r case2-prescribe}
rx2 <- prescribe_ppcs(age = 17, days_post_injury = 40, hrst = 155)
rx2
```

The BCTT-guided target of **`r rx2$target_hr` bpm** (80% of HRST) is more
precise than the age-predicted method and is the preferred approach when
equipment is available (Li, 2026, p.11).

---

## Case 3: Safety Guardrail — Vestibular Contraindication

A 15-year-old presents 30 days post-concussion but reports active dizziness
and balance difficulties.

```{r case3-screen}
screen_ppcs(
  age                 = 15,
  days_post_injury    = 30,
  vestibular_symptoms = TRUE
)
```

The package correctly routes this athlete to vestibular physiotherapy before
SSTAE can be considered. Attempting to prescribe directly also enforces the
stop rule:

```{r case3-prescribe, error = TRUE}
prescribe_ppcs(
  age                 = 15,
  days_post_injury    = 30,
  vestibular_symptoms = TRUE
)
```

---

## Case 4: Symptom Worsening — Stop-If-Worsen Rule

An athlete's PCSS increases by 3 points between sessions, triggering the
stop-if-worsen rule described in Li (2026), p.14.

```{r case4-track}
rx4 <- prescribe_ppcs(age = 16, days_post_injury = 35, hrst = 160)

# Session 1: baseline
t4a <- track_progress(
  log              = NULL,
  current_pcss     = 22,
  current_hr       = 128,
  current_duration = 20,
  prescription     = rx4
)

# Session 2: PCSS worsens by 3 points (>= 2 threshold)
t4b <- track_progress(
  log              = t4a$updated_log,
  current_pcss     = 25,
  current_hr       = 128,
  current_duration = 20,
  prescription     = rx4
)
t4b
```

The recommended HR drops by 10 bpm to **`r t4b$adjust_hr` bpm** — protecting
the athlete while keeping them in the exercise programme.

---

## Patient and Caregiver Output

For communication with patients and families, set `verbose = FALSE` for
plain-language output without clinical terminology.

```{r verbose-false}
screen_ppcs(age = 16, days_post_injury = 35, verbose = FALSE)
```

---

## Evidence Base and Limitations

PPCSexRx encodes the clinical algorithm from Li (2026), which synthesised
evidence from seven studies using GRADE methodology. Key limitations to
communicate to patients and clinicians:

- Evidence certainty is **LOW**; the recommendation is **conditional**.
- The included studies were heterogeneous in protocol and outcome measurement.
- This tool supports, but does not replace, clinical judgement.
- All prescriptions should be supervised by a qualified clinician.

## References

Li G. (2026). Sub-symptom Threshold Aerobic Exercise for Adolescents With
Persistent Post-concussion Symptoms (PPCS): A Critically Appraised Topic.
Winner, NATA Foundation Best Summary Evidence Research Award.

Leddy JJ, Haider MN, Ellis MJ, et al. Early Subthreshold Aerobic Exercise
for Sport-Related Concussion. *JAMA Pediatr.* 2019;173(4):319–325.

Kurowski BG, et al. Aerobic Exercise for Adolescents With Prolonged Symptoms
After Mild Traumatic Brain Injury. *J Head Trauma Rehabil.*
2017;32(2):79–89.

Vernau BT, Haider MN, Fleming A, et al. Exercise-Induced Vision Dysfunction
Early After Sport-Related Concussion Is Associated With Persistent
Postconcussive Symptoms. *Clin J Sport Med.* 2023;33(4):388–394.
